A general and efficient representation of Ancestral Recombination Graphs

### 2022-04-06 paper.tex

\section*{Ancestral graphs and stochastic processes}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
Hudson extended the coalescent to include
recombination~\citep{hudson1983properties,hudson1990gene}
and the model has proved to be an especially powerful basis for
simulations~\citep{hudson2002generating,baumdicker2021efficient}.
Hudson's algorithm is concerned with simulating the process of
the coalescent with recombination as efficiently as possible, and operates by
tracking segments of ancestral material carried by ancestors as
we go backwards in time. The details of Hudson's algorithm and how the
results of the ancestral process are represented are critical to later
discussions, and so we will explain the process in some detail.
See~\cite{kelleher2016efficient} for a more detailed
description of the algorithm, and how it may be implemented efficiently.

@@

Hudson's algorithm operates by tracking the state of a set of ancestral
lineages as we go backwards in time.
% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.
Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
of the form $\{(0, m, 1)\}$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount of ancestral material and
the way in which it is distributed along a chromosome.
Let $L$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (m - 1)$ where
\[
\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left- and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

% NOTE: it would probably be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
The Ancestral Recombination Graph (ARG) was later introduced by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues used the word
``ARG'' to refer to both a specific branching-coalescing stochastic process and
the graph structure derived from this process. For clarity, we will refer to the
process as the Griffiths process, or the coalescent with recombination (CwR),
and following XX-Who?-XX, we refer to the the structure as the ``big ARG'', as
opposed to the ``little ARG'' traversed by Hudson's algorithm
(\citet{wiuf1999recombination} refer to these structures as ARG and HUD respectively).

In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is now simply $k \rho$. At a recombination event
we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

The state-space of the Griffiths process is much simpler than Hudson's algorithm,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The big ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the big ARG, all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the big ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(Note that this formulation is equivalent to stipulating two
different ``types'' of node and only storing breakpoints
with recombination nodes).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

The ordering of edges is a vital element of the EARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an EARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph upwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we simply follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, which is straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left (and so we cannot emit edges as they are generated).
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

\textcolor{red}{TODO: reduce this down to one paragraph, which
defines the relationship between what we define as an ARG and
what people mean by phylogenetic networks from the literature.
We probably don't need to get into quite so much detail on
the different types that people have studies. Essentially,
we're ending the section with a brief review which situates
our definition in the literature. I think pulling apart the
idea of a ``reticulation'' vs what we call a recombinant
is essential here.}

\section*{Ancestry resolution}\label{Ancestry_resolution}
% This is a rough first pass. Will clarify terminology later.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

We can define a sample-resolved GARG as a tuple $(S, E)$
where $S \subset \mathbb{N}$ is a set of sample nodes and
$E$ is a set of edges. Each element of $E$
is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$
over which genome $c$ inherits from $p$, and there is a path from
$c$ to some non-empty subset of $S$ through the local trees.
% Should probably define how we build the local trees somewhere.

This process is illustrated in Fig~\ref{fig-ancestry-resolution}.

\section*{Evaluation of sampling probabilities}

>>>>

\section*{Ancestral graphs and stochastic processes}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
\citet{hudson1983properties,hudson1990gene} extended the coalescent to include
recombination
and the model has proved to be an especially powerful basis for
simulations~\citep{hudson2002generating,baumdicker2021efficient}.
\cite{kelleher2016efficient} provide a description of his algorithm along with
an efficient implementation, the details of which turn out to be highly relevant in 
later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).

The Ancestral Recombination Graph (ARG) was later introduced by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues used the word
``ARG'' to refer to both a specific branching-coalescing stochastic process and
the graph structure derived from this process. For clarity, we will refer to the
process as the Griffiths process, or the coalescent with recombination (CwR)
\citep{minichiello2006mapping}.

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure} if we want a simple picture
% to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities (see the ``\nameref{probabilities}'' section).

\textcolor{red}{TODO: reduce the text below down to one paragraph, which
defines the relationship between what we define as an ARG and
what people mean by phylogenetic networks from the literature.
We probably don't need to get into quite so much detail on
the different types that people have studies. Essentially,
we're ending the section with a brief review which situates
our definition in the literature. I think pulling apart the
idea of a ``reticulation'' vs what we call a recombinant
is essential here.}

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(Note that this formulation is equivalent to stipulating two
different ``types'' of node and only storing breakpoints
with recombination nodes).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}.

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph upwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we simply follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, which is straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left (and so we cannot emit edges as they are generated).
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

\section*{Ancestry resolution}\label{Ancestry_resolution}
% This is a rough first pass. Will clarify terminology later.
% Need to introduce Hudson's algorithm as concerned with simulating the process of
% the coalescent with recombination as efficiently as possible, and properly merge in
% with this section.

The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

We can define a sample-resolved gARG as a tuple $(S, E)$
where $S \subset \mathbb{N}$ is a set of sample nodes and
$E$ is a set of edges. Each element of $E$
is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$
over which genome $c$ inherits from $p$, and there is a path from
$c$ to some non-empty subset of $S$ through the local trees.
% Should probably define how we build the local trees somewhere.

% We should talk about ARG and HUD here. Fig~\ref{fig-ancestry-resolution}A shows
% ARG (the Griffiths graph). However, we have a small issue that subfig B
% doesn't strictly show HUD, because it doesn't terminate as soon as all samples have coalesced.
% We could potentially colour the top (non-HUD) section of the plot in a different colour?

This process is illustrated in Fig~\ref{fig-ancestry-resolution}. It is intimately linked to 
Hudson's algorithm which operates by
tracking segments of ancestral material carried by ancestors as
we go backwards in time.

% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.
Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
of the form $\{(0, m, 1)\}$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount of ancestral material and
the way in which it is distributed along a chromosome.
Let $L$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (m - 1)$ where
\[
\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left- and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

The state-space of the Griffiths process is much simpler than Hudson's algorithm,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The big ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the big ARG, all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the big ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''
% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

\section*{Evaluation of sampling probabilities}\label{probabilities}

@@
